Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-11 paper.qmd
Updated summarise to summarize

This is a benchmark single-cell data set collected on Human Peripheral Blood Mononuclear Cells (PBMC3k) as used in @pbmc2019. Single-cell data measures the gene expression of individual cells in a sample of tissue (see  for example, @haque2017). This type of data is used to obtain an understanding of cellular level behavior and heterogeneity in their activity. Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR often used to summarise the cluster structure. Usually, NLDR does not use the cluster labels to compute the layout, but uses color to represent the cluster labels when it is plotted.

@@

The next step is to investigate the \gD{} to understand what information is learned from this representation. @fig-model-error-mnist summarizes this investigation. Plot a shows the layout with points colored by their residual value - darker color indicates larger residual and poor fit. The plots b, c, d, e show samples of hand-written digits taken from inside the colored boxes. Going from top to bottom around the curve shape we can see that the "1"s are drawn with from right slant to a left slant. The "1"s in d (black box) tend to have the extra up stroke but are quite varied in appearance. The "1"s shown in the plots labelled e correspond to points with big residuals. They can be seen to be more strangely drawn than the others. Overall, this \gD{} layout shows a useful way to summarise the variation in way "1"s are drawn.

@@

Additional exploration of metrics to summarise the fit could be a new direction for the work. The difficulty is capturing nonlinear fits, for which Euclidean distance can be sub-optimal. We have used a very simple approach based on clustering methods, Euclidean distances to nearest centroid, which can approximate nonlinear patterns. Other cluster metrics would be natural choices to explore.

@@

Most NLDR methods only provide a \gD{} but if a \kD{} ($k>2$) layout is provided the approach developed here could be extended. Binning into cubes could be done in $3\text{-}D$ or higher, relatively easily, and used asa the basis for a wireframe of the fitted model. @barber1996 and the software @stephane2023 have algorithms for convex hulls, which \pD{} which serve as an inspiration. A simpler approach using $k$-means clustering to provide centroids could also be possible, but the complication would be to determine how to connect the centroids into an appropriate wireframe.

The new methodology is accompanied by an R package called quollr, so that it is readily usable and broadly accessible. The package has methods to fit the model, compute diagnostics and also visualise the results, with interactivity. We have primarily used the langevitour software [@harisson2024] to view the model in the data space, but other tour software such as tourr [@wickham2011] and detourr [@hart2022] could be also used.

>>>>

This is a benchmark single-cell data set collected on Human Peripheral Blood Mononuclear Cells (PBMC3k) as used in @pbmc2019. Single-cell data measures the gene expression of individual cells in a sample of tissue (see  for example, @haque2017). This type of data is used to obtain an understanding of cellular level behavior and heterogeneity in their activity. Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR often used to summarize the cluster structure. Usually, NLDR does not use the cluster labels to compute the layout, but uses color to represent the cluster labels when it is plotted.

@@

The next step is to investigate the \gD{} to understand what information is learned from this representation. @fig-model-error-mnist summarizes this investigation. Plot a shows the layout with points colored by their residual value - darker color indicates larger residual and poor fit. The plots b, c, d, e show samples of hand-written digits taken from inside the colored boxes. Going from top to bottom around the curve shape we can see that the "1"s are drawn with from right slant to a left slant. The "1"s in d (black box) tend to have the extra up stroke but are quite varied in appearance. The "1"s shown in the plots labelled e correspond to points with big residuals. They can be seen to be more strangely drawn than the others. Overall, this \gD{} layout shows a useful way to summarize the variation in way "1"s are drawn.

@@

Additional exploration of metrics to summarize the fit could be a new direction for the work. The difficulty is capturing nonlinear fits, for which Euclidean distance can be sub-optimal. We have used a very simple approach based on clustering methods, Euclidean distances to nearest centroid, which can approximate nonlinear patterns. Other cluster metrics would be natural choices to explore.

@@

Most NLDR methods only provide a \gD{} but if a \kD{} ($k>2$) layout is provided the approach developed here could be extended. Binning into cubes could be done in $3\text{-}D$ or higher, relatively easily, and used as a the basis for a wireframe of the fitted model. @barber1996 and the software @stephane2023 have algorithms for convex hulls, which \pD{} which serve as an inspiration. A simpler approach using $k$-means clustering to provide centroids could also be possible, but the complication would be to determine how to connect the centroids into an appropriate wireframe.

The new methodology is accompanied by an R package called quollr, so that it is readily usable and broadly accessible. The package has methods to fit the model, compute diagnostics and also visualize the results, with interactivity. We have primarily used the langevitour software [@harisson2024] to view the model in the data space, but other tour software such as tourr [@wickham2011] and detourr [@hart2022] could be also used.
